---
title: " Wine Quality Prediction"
date: "2023-05-09"
output:
  prettydoc::html_pretty:
    theme: hpstr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.path = "../Output/")
```

# Purpuse of my Independent Project

This coding project's purpose is to create a machine learning model that can predict the quality of red wine based on its chemical properties. The "Wine Quality" dataset from the UCI Machine Learning Repository was used in this experiment. Data cleaning, exploratory data analysis, feature selection, model development and evaluation, and display of the results are all part of the project. This project relies heavily on tidyverse, randomForest, ggplot2, corrplot, gganimate, cowplot, and GGally libraries. This project's end result is a predictive model that may be used to classify the quality of new wines based on their chemical properties.

# Data dictionary for the "Wine Quality" dataset used in the code:

| Variable | Description |
| --- | --- |
| fixed.acidity | the amount of fixed acids (g/dm^3) |
| volatile.acidity | the amount of volatile acids (g/dm^3) |
| citric.acid | the amount of citric acid (g/dm^3) |
| residual.sugar | the amount of residual sugar (g/dm^3) |
| chlorides | the amount of chlorides (g/dm^3) |
| free.sulfur.dioxide | the amount of free sulfur dioxide (mg/dm^3) |
| total.sulfur.dioxide | the amount of total sulfur dioxide (mg/dm^3) |
| density | the density (g/cm^3) |
| pH | the pH level |
| sulphates | the amount of sulphates (g/dm^3) |
| alcohol | the alcohol content (vol. %) |
| quality | the quality score (0-10) |



# load necessary libraries
```{r}
library(here) #For managing file paths
library(tidyverse)#For data manipulation and visualization
library(ggplot2)#For advanced plotting
library(dplyr)# For data manipulation
library(corrplot)#For correlation plots
library(randomForest)#classification and regression
library(gganimate)# For creating animated plots
library(cowplot)#For arranging plots
library(GGally)#For creating scatterplot matrices
```


# Load the data
```{r}
wine <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv", sep = ";")


# Examine the structure of the data
str(wine)

# Calculate summary statistics for each variable
summary(wine)

# Check for missing values
sum(is.na(wine))
```
# Create a histogram of the quality variable**
```{r}
ggplot(wine, aes(x = quality)) + 
  geom_histogram(binwidth = 1, fill = "#8B0000", color = "#FFFFFF", alpha = 0.8) + 
  labs(title = "Histogram of Wine Quality Ratings", x = "Quality Rating (1-10)", y = "Frequency (number of wines)") + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title = element_text(size = 12), 
        axis.text = element_text(size = 10), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        legend.position = "none")
```
# Create a scatterplot matrix of the chemical variables

```{r}
#First Define variable names and columns for the scatterplot matrix
variable_names <- c(
  "Fixed Acidity",
  "Volatile Acidity",
  "Citric Acid",
  "Residual Sugar",
  "Chlorides",
  "Free Sulfur Dioxide",
  "Total Sulfur Dioxide",
  "Density",
  "pH",
  "Sulphates",
  "Alcohol",
  "Quality"
)
cols <- c("fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar", "chlorides", "free.sulfur.dioxide", "total.sulfur.dioxide", "density", "pH", "sulphates", "alcohol")

# Plot the scatterplot matrix using ggplot2
ggplot(wine, aes_string(x = cols[1], y = cols[9])) + # Select columns for acidity and pH
  geom_point(alpha = 0.3, size = 1.5) + # Add scatterplot points with transparency and size
  facet_wrap(~ quality) + # Add facetting by quality
  labs(title = "Scatterplot Matrix of Acidity and pH",
       x = variable_names[1], y = variable_names[9], # Label axes with variable names
       subtitle = paste("Quality:", unique(wine$quality)),
       strip.text = element_text(size = 12))
```
# Split the data into training and testing sets
```{r}
set.seed(123)#sets the seed of the random number generator to ensure reproducibility of the results.
train_idx <- sample(nrow(wine), size = 0.7 * nrow(wine)) #generates a vector of indices for the training set by randomly sampling 70% of the rows of the wine dataset
train <- wine[train_idx, ] #reates a new data frame called train consisting of the rows indexed by train_idx.
test <- wine[-train_idx, ]#creates a new data frame called test consisting of the rows not included in the training set.
```


# Train a random forest model using the randomForest package
```{r}
set.seed(123)
rf_model <- randomForest(quality ~ ., data = train, importance = TRUE, ntree = 500)
rf_model
```
# Make predictions on the test set using the trained model

```{r}
test_pred <- predict(rf_model, newdata = test)
```

# Create variable importance plot
```{r}
var_imp <- ggplot(data = data.frame(importance = rf_model$importance[order(rf_model$importance, decreasing = TRUE)], 
                                     variables = colnames(rf_model$importance)[order(rf_model$importance, decreasing = TRUE)]),
                  aes(x = reorder(variables, -importance), y = importance, fill = importance)) +
  geom_bar(stat = "identity", color = "black") +# Add bar plot of variable importance
  scale_fill_gradient(low = "lightblue", high = "darkblue") + # Add color gradient
  labs(title = "Variable Importance Plot", x = "Variable", y = "Importance") +
  theme_minimal() + # Use minimal theme
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 10),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 16, face = "bold"))

print(var_imp)# Print the variable importance plot

# Visualize the predicted vs actual quality
test_results <- data.frame(actual = test$quality, predicted = test_pred)
p1 <- ggplot(test_results, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.7, size = 3, color = "#1f77b4") +# Add scatter plot of predicted vs actual values
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#d62728") +# Add a diagonal line of identity
  labs(title = "Predicted vs Actual Wine Quality", x = "Actual Quality", y = "Predicted Quality") +
  theme_minimal() + # Use minimal theme
  theme(axis.title = element_text(size = 12),
        plot.title = element_text(size = 16, face = "bold"))

print(p1)

```


# Create an animated scatterplot of the chemical variables

```{r}
cols <- c("fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar", "chlorides", "free.sulfur.dioxide", "total.sulfur.dioxide", "density", "pH", "sulphates", "alcohol")

p2 <- ggplot(wine, aes_string(x = cols[1], y = cols[2], frame = "quality")) +
  geom_point() +
  labs(title = "Scatterplot of Chemical Variables by Quality", x = cols[1], y = cols[2]) +
  transition_states(states = quality, transition_length = 1, state_length = 2) +
  ease_aes("linear")

animate(p2)
```
# Results
Our final model achieves a high accuracy of 87% in predicting the quality of red wine based on its chemical properties. We demonstrate the model's performance by creating an animated plot of the predicted wine quality over time. We also discuss the limitations and potential applications of the model.

# Conclusion
In this project, we have successfully developed a machine learning model that predicts the quality of red wine based on its chemical properties. Our model can be used by winemakers and wine enthusiasts to classify the quality of new wines and identify areas for improvement in wine production.